---
title: "Embedding Visualizations"
author: "Vivek Sriram"
date: '2023-06-07'
output: pdf_document
---

```{r}
library(ggplot2)
library(dplyr)
library(umap)
library(gridExtra)
library(ggpubr)
```

# Read in input data files

Includes k-means clustering results with k=5
```{r}
embeddingMat_m = read.csv("/Users/viveksriram/Desktop/embeddingMat_m_fixed.csv")
embeddingMat_m$kmeansLabel = as.factor(embeddingMat_m$kmeansLabel)

embeddingMat_f = read.csv("/Users/viveksriram/Desktop/embeddingMat_f_fixed.csv")
embeddingMat_f$kmeansLabel = as.factor(embeddingMat_f$kmeansLabel)
```

H53-H54 (visual disturbances and blindness)
H10-H13 (disorders of conjunctiva)
H25-H28 (disorders of lens)

```{r}
interestingDiseases = c("C64-C68", "F20-F29", "I20-I25", "L40-L45", "M05-M14", "M15-M19", "M45-M49")

embeddingMat_f$toBeLabeled = ifelse(embeddingMat_f$Trait %in% interestingDiseases, 1, 0)
embeddingMat_m$toBeLabeled = ifelse(embeddingMat_m$Trait %in% interestingDiseases, 1, 0)
```

Female
M05 - M14: Inflammatory polyathropathies (rheumatoid arthritis)
M15 - M19: Osteoarthritis
M80 - M85: Osteoporosis
L80 - L99: Scleroderma
G35 - G37: Multiple sclerosis

Male
M45 - M49: Ankylosing spondylitis
I20 - I25: Ischemic heart diseases
C43 - C44: Melanoma
E70 - E90: Metabolic disorders
L40 - L45: Psoriasis


# PCA
```{r}
pca_category_m = ggplot(embeddingMat_m, 
       aes(x=pca.2d.one, 
           y=pca.2d.two,
           color=Disease.Category,
           label=Trait)) +
  geom_point() +
  geom_label(aes(pca.2d.one, pca.2d.two, label=Trait), 
            data = embeddingMat_m[embeddingMat_m$Trait %in% interestingDiseases,]) +
  labs(title="Male DDN",
       x="Principal Component 1", 
       y = "Principal Component 2") +
  xlim(-0.28, 0.28) +
  ylim(-0.28, 0.28)
```

```{r}
pca_category_f = ggplot(embeddingMat_f, 
       aes(x=pca.2d.one, 
           y=pca.2d.two,
           color=Disease.Category)) +
  geom_point() +
  geom_label(aes(pca.2d.one, pca.2d.two, label=Trait), 
            data = embeddingMat_f[embeddingMat_f$Trait %in% interestingDiseases,]) +
  labs(title="Female DDN",
       x="Principal Component 1", 
       y = "Principal Component 2") +
  xlim(-0.28, 0.28) +
  ylim(-0.28, 0.28)
```

```{r}
ggarrange(pca_category_f, pca_category_m, 
          common.legend = TRUE)
```

```{r}
pca_kmeans_m = ggplot(embeddingMat_m, 
       aes(x=pca.2d.one, 
           y=pca.2d.two,
           color=kmeansLabel)) +
  geom_point() +
  geom_label(aes(pca.2d.one, pca.2d.two, label=Trait), 
            data = embeddingMat_m[embeddingMat_m$Trait %in% interestingDiseases,]) +
  labs(title="Male DDN",
       x="Principal Component 1", 
       y = "Principal Component 2") +
  xlim(-0.28, 0.28) +
  ylim(-0.28, 0.28)
```

```{r}
pca_kmeans_f = ggplot(embeddingMat_f, 
       aes(x=pca.2d.one, 
           y=pca.2d.two,
           color=kmeansLabel)) +
  geom_point() +
  geom_label(aes(pca.2d.one, pca.2d.two, label=Trait), 
            data = embeddingMat_f[embeddingMat_f$Trait %in% interestingDiseases,]) +
  labs(title="Female DDN",
       x="Principal Component 1", 
       y = "Principal Component 2") +
  xlim(-0.28, 0.28) +
  ylim(-0.28, 0.28)
```

```{r}
ggarrange(pca_kmeans_f, pca_kmeans_m, common.legend = TRUE)
```


# UMAP
```{r}
embeddingData_f = embeddingMat_f %>%
  select(X0, X1, X2, X3, X4, X5, X6, X7, X8, X9, X10, X11, X12, X13, X14, X15)

embeddingData_m = embeddingMat_m %>%
  select(X0, X1, X2, X3, X4, X5, X6, X7, X8, X9, X10, X11, X12, X13, X14, X15)
```

```{r}
embeddingDataM.umap <- umap(embeddingData_m, n_neighbors = 3)
embeddingDataF.umap <- umap(embeddingData_f, n_neighbors = 3)
```

```{r}
embeddingMat_m$umapDim1 = embeddingDataM.umap$layout[,1]
embeddingMat_m$umapDim2 = embeddingDataM.umap$layout[,2]

embeddingMat_f$umapDim1 = embeddingDataF.umap$layout[,1]
embeddingMat_f$umapDim2 = embeddingDataF.umap$layout[,2]
```



```{r}
umap_category_f = ggplot(embeddingMat_f, 
       aes(x=umapDim1, 
           y=umapDim2,
           color=Disease.Category)) +
  geom_point() +
  geom_label(aes(umapDim1, umapDim2, label=Trait), 
            data = embeddingMat_f[embeddingMat_f$Trait %in% interestingDiseases,]) +
  labs(title="Female DDN",
       x="UMAP Dimension 1", 
       y = "UMAP Dimension 2") +
  xlim(-5, 5) +
  ylim(-5,5)
```

```{r}
umap_category_m = ggplot(embeddingMat_m, 
       aes(x=umapDim1, 
           y=umapDim2,
           color=Disease.Category,
           label=Trait)) +
  geom_point() +
  geom_label(aes(umapDim1, umapDim2, label=Trait), 
            data = embeddingMat_m[embeddingMat_m$Trait %in% interestingDiseases,]) +
  labs(title="Male DDN",
       x="UMAP Dimension 1", 
       y = "UMAP Dimension 2") +
  xlim(-5, 5) +
  ylim(-5,5)
```

```{r}
ggarrange(umap_category_f, umap_category_m, common.legend = TRUE)
```



```{r}
umap_kmeans_f = ggplot(embeddingMat_f, 
       aes(x=umapDim1, 
           y=umapDim2,
           color=kmeansLabel)) +
  geom_point() +
  geom_label(aes(umapDim1, umapDim2, label=Trait), 
            data = embeddingMat_f[embeddingMat_f$Trait %in% interestingDiseases,]) +
  labs(title="Female DDN",
       x="UMAP Dimension 1", 
       y = "UMAP Dimension 2") +
  xlim(-5, 5) +
  ylim(-5,5)
```

```{r}
umap_kmeans_m = ggplot(embeddingMat_m, 
       aes(x=umapDim1, 
           y=umapDim2,
           color=kmeansLabel)) +
  geom_point() +
  geom_label(aes(umapDim1, umapDim2, label=Trait), 
            data = embeddingMat_m[embeddingMat_m$Trait %in% interestingDiseases,]) +
  labs(title="Male DDN",
       x="UMAP Dimension 1", 
       y = "UMAP Dimension 2") +
  xlim(-5, 5) +
  ylim(-5,5)
```

```{r}
ggarrange(umap_kmeans_f, umap_kmeans_m, common.legend = TRUE)
```





```{r}
traitDescriptions = read.csv("/Users/viveksrm/Desktop/traitDescription_diseaseCategories.csv") %>%
  select(Trait, Description, Disease.Category)
```

# ssDDN 1e-4 

## p = 1, q = 1, 100 runs, 10000 walks, random seed starting at 1
```{r}
# Embedding distances for individual diseases
embeddingDistances = read.csv("/Users/viveksrm/Desktop/embeddingDistances_neg4_100runs_10000walks.csv", sep = "\t")

embeddingDistances = left_join(embeddingDistances, traitDescriptions, by = "Trait")

embeddingDistances_sorted <- embeddingDistances[order(embeddingDistances$meanDist),] 
```


```{r}
# Embedding distances for individual diseases
categoryDistances = read.csv("/Users/viveksrm/Desktop/embeddingDistances_neg4_100runs_10000walks.csv", sep = "\t")

categoryDistances_sorted <- categoryDistances[order(categoryDistances$meanDist),] 
```



```{r}
head(embeddingDistances_sorted)
```
Cosine lowest 5 distances
- N20 to N23 Urolithiasis
- K65 to K67 Diseases of peritoeneum
- H30 to H36 Disorders of choroid and retina
- M91 to M94 Chondropathies
- M15 to M19 Arthrosis

```{r}
tail(embeddingDistances_sorted)
```
Cosine highest 5 distances:
- I20 to I25 Ischaemic heart diseases
- N25 to N29 Other disorders of kidney and ureter
- H00 to H06 Disorders of eyelid, lacrimal system and orbit
- L80 to L99 Other disorders of the skin and subcutaneous tissue
- H25 to H28 Disorders of lens




```{r}
p<-ggplot(embeddingDistances_sorted[72:102,], aes(x=meanDist, y= reorder(Trait, meanDist))) + 
  geom_point()+
  geom_errorbar(aes(xmin=meanDist-stdDevDist, xmax=meanDist+stdDevDist), width=.2,
                position=position_dodge(0.05))
  
p
```

```{r}
p<-ggplot(embeddingDistances_sorted[1:30,], aes(x=meanDist, y= reorder(Trait, meanDist))) + 
  geom_point()+
  geom_errorbar(aes(xmin=meanDist-stdDevDist, xmax=meanDist+stdDevDist), width=.2,
                position=position_dodge(0.05))
  
p
```

```{r}
range(embeddingDistances_neg4_p1q1_sorted$meanDist)
median(embeddingDistances_neg4_p1q1_sorted$meanDist)
mean(embeddingDistances_neg4_p1q1_sorted$meanDist)
```

```{r}
ggplot(embeddingDistances_neg4_p1q1_sorted, aes(x=meanDist)) + geom_histogram()
```



```{r}
p<-ggplot(embeddingDistances_neg4_p1q1_sorted[72:102,], aes(x=meanDist, y= reorder(Trait, meanDist))) + 
  geom_point()+
  geom_errorbar(aes(xmin=meanDist-stdDevDist, xmax=meanDist+stdDevDist), width=.2,
                position=position_dodge(0.05))
  
p
```

```{r}
p<-ggplot(embeddingDistances_neg4_p1q1_sorted[1:30,], aes(x=meanDist, y= reorder(Trait, meanDist))) + 
  geom_point()+
  geom_errorbar(aes(xmin=meanDist-stdDevDist, xmax=meanDist+stdDevDist), width=.2,
                position=position_dodge(0.05))
  
p
```

```{r}
range(embeddingDistances_neg4_p1q1_sorted$meanDist)
median(embeddingDistances_neg4_p1q1_sorted$meanDist)
mean(embeddingDistances_neg4_p1q1_sorted$meanDist)
```

```{r}
ggplot(embeddingDistances_neg4_p1q1_sorted, aes(x=meanDist)) + geom_histogram()
```


